---
title: "Practical Machine Learning Assignment"
author: "Mateusz Pawlowski"
date: "24th of May 2016"
output: html_document
---

#Introduction

This report is intended to document the final assignment in the Practical Machine Learning course provided by the Coursera platform. 

##Project Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


The training data for this project are available
[here:](https://d396qusza40orc.cloudf.trainingront.net/predmachlearn/pml-training.csv)

The test data are available
[here:](https://d396qusza40orc.cloudf.trainingront.net/predmachlearn/pml-testing.csv)


#Goal of the analysis 

The goal of the  “classe” variable in the training set, which represents the manner of the excercise.  

#Getting the data

```{r setup}
#Basic info on the session
sessionInfo()
#Loading packages
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
library(dplyr)

#Downloading data
training.d <- read.csv(url("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"), na.strings=c("NA","#DIV/0!",""))

testing.d  <- read.csv(url("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"), na.strings=c("NA","#DIV/0!",""))
```


```{r}
#Split the dataset into the testing and training dataframes
toTrain = createDataPartition(training.d$classe, p = 0.75)[[1]]
training = training.d[toTrain,]     # select 75% of records
testing = training.d[-toTrain,]     # keep 25% of the rest
#Lets get a glimpse of the data
kable((glimpse(training)))
```

#Cleaning data

```{r}
#The first 6 columns carry unnecessary information
df.training <- training[,7:160]
df.testing <- testing[,7:160]
## Find columns with near zero variance and remove them
nzv_cols <- nearZeroVar(df.training)
if(length(nzv_cols) > 0) df.training <- df.training[, -nzv_cols]
# Remove Columns containing missing values
df.training <- df.training[ , colSums(is.na(df.training)) == 0]
```

#Let's build three models and compare their accuracy. 1st model is constructed using a decision tree technique

```{r}
set.seed(123456)
dtree.model <- rpart(classe ~ ., data=df.training, method="class")
fancyRpartPlot(dtree.model)
pred.dtree <- predict(dtree.model, df.testing, type = "class")
cmtree <- confusionMatrix(pred.dtree, df.testing$classe)
cmtree
```

#A random forest model

```{r}
set.seed(12345)
rfmodel <- randomForest(classe ~ ., data=df.training)
pred.randomf <- predict(rfmodel, df.testing, type = "class")
plot(rfmodel)
cmrf <- confusionMatrix(pred.randomf, df.testing$classe)
cmrf
```

#GBM model 

```{r}
set.seed(123456)
t.control <- trainControl(method = "repeatedcv",
                           number = 5,
                           repeats = 1)

gbmod <- train(classe ~ ., data=df.training, method = "gbm",
                 trControl = t.control ,
                 verbose = FALSE)

gbmFinMod1 <- gbmod$finalModel

gbmTest <- predict(gbmod, newdata=df.testing)
gbmAccuracyTest <- confusionMatrix(gbmTest, df.testing$classe)
gbmAccuracyTest
```

#Predicting the results

As the random forest model achieved the highest level of accuracy we use it to predict the values in the testing set.

```{r}
predictionRFmodel <- predict(rfmodel, testing.d, type = "class")
predictionRFmodel
```

